Article Title

### 2014-01-20 plos.tex
rev37. Responded to all comments.

\DeclareMathOperator{\Tr}{tr}
\newcommand{\sq}[1]{\lq#1\rq}
\newcommand{\mcond}{\,\middle\vert\,}
\newcommand{\cond}{\,\vert\,}
\newcommand{\figref}[2]{Fig.\;\ref{fig:#1}\,#2}
\newcommand{\loss}[1]{\mathcal L\left(#1\right)} 
\newcommand{\eloss}[1]{\mathcal L_0\left(#1\right)}
\newcommand{\T}{{\sf T}}
\newcommand{\E}[2][]{\mathbb E_{#1}\left[ #2\right]}    % expected value

@@

\section*{Abstract}
% Please keep the abstract between 250 and 300 words
Ambitious projects aim to record the activity of ever larger and denser subsets of neurons \emph{in vivo}.  Correlations in neural activity measured in such recordings can reveal important aspects of the organization of neural circuits.  However, estimating and interpreting large correlation matrices is challenging.  Estimation can be improved by regularization, \emph{i.e.}\;by imposing a structure on the estimate.  The amount of improvement depends on how closely the assumed structure represents dependencies in the data. Therefore, the selection of the most efficient correlation matrix estimator for a given neural circuit must be determined empirically.  Importantly, the identity and structure of the most efficient estimator informs about the types of dominant dependencies governing the system.

@@

Pearson correlations between the spiking activity of pairs of neurons, are among the most familiar measures of functional connectivity \cite{Averbeck:2006, Zohary:1994, Kohn:2005, Bair:2001, Ecker:2010}.  In particular, \emph{noise correlations}, \emph{i.e.}\;the correlations of trial-to-trial response variability between pairs of neurons, have a profound impact on stimulus coding \cite{Zohary:1994, Abbott:1999, Sompolinsky:2001, Nirenberg:2003, Averbeck:2006, Josic:2009, Berens:2011, Ecker:2011}. In addition, noise correlations and correlations in spontaneous activity have been hypothesized to reflect key features of the anatomical connectivity \cite{Gerstein:1964}.  This interpretation is supported by a series of discoveries of nontrivial relationships between neural correlations and other aspects of circuit organization such as the physical distances between neurons \cite{Smith:2008, Denman:2013}, their synaptic connectivity \cite{Ko:2011},  stimulus response similarity \cite{Bair:2001, Arieli:1995, Chiu:2002, Kenet:2003, Kohn:2005, Cohen:2008, Cohen:2009, Ecker:2010, Rothschild:2010, Ko:2011, Smith:2013b}, cell types \cite{Hofer:2011}, cortical layer specificity \cite{Hansen:2012, Smith:2013}, progressive changes in development and in learning \cite{Golshani:2009, Gu:2011}, changes due to sensory stimulation and global brain states \cite{Goard:2009, Kohn:2009, Ecker:2010, Renart:2010} among others.

Neural correlations do not come with ready or unambiguous mechanistic interpretations. They can arise from monosynaptic or polysynaptic interactions, common or correlated inputs, oscillations, top-down modulation, and background network fluctuations \cite{Perkel:1967, Moore:1970, Shadlen:1998, Salinas:2001, Ostojic:2009, Rosenbaum:2011}. But multineuronal recordings do provide more information than available from an equivalent number of pairs of cells considered separately. For example, the eigenvalue decomposition of the covariance matrix expresses shared correlated activity components across the population; common fluctuations of population activity may be accurately represented by just a few principal components but will affect all correlation coefficients. In contrast, the off-diagonal elements of the precision matrix (inverse of the covariance) constitute scaled partial correlations between neuron pairs.  The partial correlation coefficient between two neurons reflects their linear association conditioned on the activity of all the other recorded cells.  For this reason, partial correlations have been proposed to more closely reflect the direct effects between components of complex systems and have been used to describe gene interactions \cite{Schafer:2005, Peng:2009} and the interactions between brain regions \cite{Varoquaux:2012, Ryali:2012}.

However, estimation of covariance matrices from large populations presents a number of numerical challenges:  The amount of recorded data grows only linearly with population size whereas the number of estimated coefficients increases quadratically.  This mismatch leads to an increase in spurious correlations, overestimation of common activity (\emph{i.e.}\;overestimation of large eigenvalues) \cite{Ledoit:2004}, and poorly conditioned partial correlations \cite{Schafer:2005}. 
The \emph{sample correlation matrix}, which is an unbiased estimate, is often used to estimate the true correlation matrix (ref). However, it has many free parameters, and is thus highly sensitive to sampling noise. As a result, on average, the sample correlation matrix is farther from the true correlation matrix than some structured estimates.

@@

To address these questions, in addition to the sample covariance matrix, we evaluated four regularized covariance matrix estimators that imposed different kinds of structure on the estimate. The estimators are designated as follows:
\begin{description}
\item[$C_{\sf 0}$] -- sample covariance matrix, the unbiased, unregularized estimator.
\item[$C_{\sf diag}$] -- linear shrinkage toward a diagonal covariance matrix. 
\item[$C_{\sf factor}$] -- linear shrinkage toward a low-rank matrix produced by factor analysis.
\item[$C_{\sf sparse}$] --  sparse partial correlations, \emph{i.e.}\;covariance selection.
\item[$C_{\sf sparse+latent}$] -- sparse partial correlations between the recorded neurons combined with partial correlations with several inferred latent units.
\end{description}

@@

\section*{Results}
% Results and Discussion can be combined.
\paragraph{Covariance estimation}
The covariance matrix defined as
\begin{equation}\label{eq:true-covariance}
    \Sigma = \E{(x-\mu)(x-\mu)^\T},\quad \mu = \E{x}
\end{equation}
where $\E{\cdot}$ denotes expectation; $x$ is the $p\times 1$ vector of real-valued instantaneous firing rates in bins of duration $\Delta t$; and $\mu$ is the vector of mean firing rates.  Usually, neural covariance matrices have been estimated as the \emph{sample covariance matrix} $C_{\sf 0}$ calculated from the empirical sample of observations $x(t),\; t=n(1,\ldots,n)$ as
\begin{equation}
    C_{\sf 0} = \frac 1 \nu \sum\limits_{t=1}^n (x(t)-\bar x)(x(t)-\bar x)^\T,\quad \bar x= \frac 1 n \sum\limits_{t=1}^n x(t)
\end{equation}
where $\nu$ is the number of degrees of freedom per neuron in the sample ($\nu=n-1$ if observations are independent). In this study, we estimate the true mean, $\mu$, using the sample mean $\bar x$, but seek a better estimate of the covariance matrix than $C_{\sf 0}$.  In the case of noise correlations, the true mean $\mu$ and the sample mean $\bar x$ are conditioned on the stimulus.

Given a covariance matrix estimate $C$ (not necessarily the sample covariance $C_{\sf 0})$), the correlation matrix $R$ is calculated by normalizing $C$ by its diagonal (variance estimates):
\begin{equation}\label{eq:precision}
    R = \left(I\circ C\right)^{-\frac 1 2} C \left(I\circ C\right)^{-\frac 1 2},
\end{equation}
where $\circ$ denotes entrywise matrix product (Hadamard product) and $I$ is the $p\times p$ identity matrix. 
Similarly, the matrix of partial correlations $P$ is computed by normalizing the \emph{precision matrix} $C^{-1}$ (inverse of the covariance matrix):
\begin{equation}\label{eq:partial}
    P = -\left(I\circ C^{-1}\right)^{-\frac 1 2} C^{-1} \left(I\circ C^{-1}\right)^{-\frac 1 2}
\end{equation}

We considered four regularized estimators based on distinct families of low-dimensional target estimates: $C_{\sf diag}$, $C_{\sf factor}$, $C_{\sf sparse}$, and $C_{\sf sparse+latent}$. In probabilistic models with exclusively linear dependencies, their targets estimates can be described by distinct families of graphical models (\figref{02}{\,Row}{row 1}).

\begin{FPfigure}
    \begin{center}
        \includegraphics{./figures/Figure02.pdf}
    \end{center}
    \caption{{\bf Estimators whose regularization targets represent the structure of the true covariance matrix outperform other estimators.}
        {\bf Row 1.} Graphical representations of the regularization targets in the four regularized covariance matrix estimators used in this study,   structures of interactions between observed neurons (green spheres) and latent units (light-shaded spheres).
        In $C_{\sf diag}$, the regularization target is  \sq{diagonal} ({\bf A}), observed neurons exert no linear effects on one another neither directly nor through interactions with common latent units. 
        In \sq{latent factors} ({\bf B}), the correlated activity of observed cells is driven by several latent units. 
        In \sq{sparse interactions} ({\bf C}), the correlation matrix is defined by a set of linear interactions between observed neurons. 
        In \sq{sparse+latent} ({\bf D}), correlations arise through direct linear interactions in a sparse subset of observed neurons and through interactions with common latent units. 
        {\bf Row 2.} Examples of $50\times 50$ correlation matrices corresponding to each type of low-dimensional structure. 
        The factor model ({\bf B}) has three latent units. 
        The partial correlation matrix of the sparse model ({\bf C}) is 73\% sparse.
        The \sq{sparse+latent} matrix has one latent unit and its direct interactions are 78\% sparse.
        {\bf Row 3.} Examples of sample correlation matrices calculated from samples of 1000 observations taken from simulated random processes with corresponding correlation matrices from row 2.
        {\bf Row 4.} Mean \emph{excess loss} (Eq.~\ref{eq:excess-loss}) attained by each of the five estimators as a function of sample size. The error bars indicate the standard error of the mean based on 30 samples.
        {\bf Row 5.} Mean \emph{cross-validation loss} (Eq.~\ref{eq:rel-cv-loss}) of covariance estimators with respect to the matching estimator. The values are relative to the validation loss of the estimator that matches the low-dimensional structure of the true covariance matrix. Error bars indicate the standard error of the mean based on 30 samples.
    }
    \label{fig:02}
\end{FPfigure}

For estimator $C_{\sf diag}$, the target estimate is the diagonal matrix $D$ whose diagonal elements contain the variance estimates. Regularization is achieved by linear \emph{shrinkage} of the unbiased estimate $C_{\sf 0}$ toward $D$ controlled by the scalar \emph{shrinkage intensity} parameter $\lambda \in [0, 1]$:
\begin{equation}\label{eq:c-diag}
C_{\sf diag} = (1-\lambda) C_{\sf 0} + \lambda D
\end{equation}
The structure of target $D$ favors the absence of linear associations between the activity of observed neurons (\figref{02}{A Row 1}).  If sample correlations arise from spurious or contingent effects, $C_{\sf diag}$ should outperform the other estimators.

The target of estimator $C_{\sf factor}$ is the factor model $L + \Psi$, where $L$ is the $p\times p$ positive definite matrix of rank $d$, and $\Psi$ is the diagonal matrix of independent variances.  The estimator, 
\begin{equation}\label{eq:c-factor}
C_{\sf factor} = (1-\lambda) C_{\sf 0} + \lambda (L + \Psi),
\end{equation}
has two hyperparameters: the number of factors $d$ and shrinkage intensity $\lambda$. The target estimate $L + \Psi$ has the structure that arises when correlated fluctuations in population activity are linearly driven by a number of latent factors that affect many cells while direct interactions between cells are insignificant (\figref{02}{B}).

Estimator $C_{\sf sparse}$ favors correlations that arise from direct linear effects between pairs of observed neurons when only a fraction of neuron pairs interact (\figref{02}{C}).  This assumption is enforced by fitting only a subset of the off-diagonal elements of the precision matrix while setting the rest to zero. The estimator takes the form
\begin{equation}\label{eq:c-sparse}
C_{\sf sparse} = S^{-1},
\end{equation}
where $S$ is a sparse matrix with a large fraction of zeros in its off-diagonal.  The estimate has one hyperparameter that determines the sparsity (fraction of off-diagonal zeros) of $S$.

Estimator $C_{\sf sparse+latent}$ allows the assumptions of both $C_{\sf factor}$ and $C_{\sf sparse}$.  Its correlation structure can be explained by linear effects between some pairs of observed neurons and by several latent units (\figref{02}{D}). The estimator has the form \cite{Chandrasekaran:2010,Ma:2013} (See Methods):
\begin{equation}\label{eq:c-sl}
C_{\sf sparse+latent} = (S - L)^{-1}
\end{equation}
where, as above, $S$ is a sparse matrix and $L$ is a $p\times p$ matrix of low rank $d$. The estimator has two hyperparameters: one to regulate the rank $d$ and the other to regulate the sparsity of matrix $S$.

\paragraph{Simulation}
To illustrate the performance of the four regularized estimators, we constructed four model populations of $p=50$~neurons each with each matching the correlation structures of a different target estimate (\figref{02}{\,Row 2}).  Sample correlation matrices calculated from samples of size $n=1000$ contain substantial sampling noise (\figref{02}{\,Row 3}) that makes the correlation structure less obvious.

An evaluation of the performance of a covariance matrix estimator $C$ requires a real-valued \emph{loss function} $\loss{C,\Sigma}$. The loss function quantifies the deviation of the estimate $C$ from truth $\Sigma$ and attains its minimum when $C=\Sigma$.

In this study, we adopted the \emph{normal loss} defined as
\begin{equation}\label{eq:loss}
    \loss{C,\Sigma} = \frac 1 p\left[\ln \det C + \Tr(C^{-1}\Sigma)\right]
\end{equation}

This loss function relates to the multivariate normal log-likelihood function $L(\Sigma|C_{\sf 0})$ through the following expression:
\begin{equation}
    L(\Sigma|C_{\sf 0}) = -\frac p 2 \ln(2\pi) -\frac p 2 \loss{\Sigma,C_{\sf 0}}
\end{equation}
Although the two functions appear similar in form, they differ conceptually: The log-likelihood is a function of the parameter $\Sigma$ given the sample covariance matrix $C_{\sf 0}$ whereas the loss function expresses the distance from an arbitrary estimate $C$ to the truth $\Sigma$.

The normalization by $p$ makes the values of the loss function comparable across different population sizes.  Therefore, the loss function is measured in nats per neuron per observation (or frame).

The choice of loss function is motivated by mathematical convenience. We expect that the main conclusions of our study will not change qualitatively with other well behaved loss functions, such as the Frobenius norm of the difference \cite{Ledoit:2004,Schafer:2005}, Stein's entropy loss, or quadratic loss \cite{James:1961,Fan:2008}.

We define the \emph{excess loss} as
\begin{equation}\label{eq:excess-loss}
    \eloss{C,\Sigma} = \loss{C,\Sigma}-\loss{\Sigma,\Sigma},
\end{equation}
which assumes zero at its minimum.

We computed the excess losses for 30 samples for each sample size $n=250$, 500, 10000, 2000, and 4000 for each of the five estimators and for each ground truth (\figref{02}{\,row 5}). For each estimator, hyperparameters were estimated by nested cross-validation (See Methods.).  As expected, estimators whose structure matched that of the true model typically outperformed the other estimators. There were two exceptions: First, for some smaller samples, the data were insufficient to reveal the true correlation structure and the estimators with the simpler structure, $C_{\sf diag}$ and $C_{\sf factor}$ outperformed the estimators with matching structure (\figref{02}{\,columns C and D, rows 5 and 6}).  Second, $C_{\sf sparse+latent}$ often performed equally well to the sparse estimator even when the ground truth did not have any latent units.  In these cases, it inferred the right number of latent units, zero, and estimated optimally.

When the ground truth, $\Sigma$, is not accessible, loss can be estimated solely from the data through \emph{validation}.  In validation, an additional independent \emph{testing sample} is used to compute a sample covariance estimate $C_{\sf 0}^\prime$.  Then \emph{validation loss} $\loss{C,C_{\sf 0}^\prime}$ can be used as an estimate of loss $\loss{C,\Sigma}$.  Thus, estimators resulting in consistently lower validation loss can be inferred to produce estimates that are closer to truth than estimators with higher validation loss.

@@

Then the cross-validation loss is an unbiased estimate of loss:
\begin{equation*}
    \E[C_{\sf 0}^\prime]{\loss{C,C_{\sf 0}^\prime}}=\loss{C,\E[C_{\sf 0}^\prime]{C_{\sf 0}^\prime}}=\loss{C,\Sigma}
\end{equation*}

The property of additivity does not hold for other popular loss functions such as Stein's entropy loss or various quadratic losses.  With non-additive loss functions, validation loss is a biased estimate of loss, but it is still generally appropriate to be used in lieu of loss.

With empirical data, acquiring a separate testing sample is not practical. Instead, $K$-fold cross-validation is used: The sample is divided into $K$ subsets of approximately equal size ($K=10$ in this study).  Then each subset is used as the validation sample with the other $K-1$ serving as the training sample. The validation losses from each of such \sq{folds} are averaged to produce the \emph{cross-validation loss} or \emph{CV-loss} for short.  Let $C_{\sf 0}^{\{k\}}$ denote the sample covariance matrix computed from the $k^{th}$ subset and $C^{\{\setminus k\}}$ denote the results of estimator $C$ trained on the remaining $K-1$ subsets. Then cross-validation loss for estimator $C$ is
\begin{equation}\label{eq:cv-loss}
    \ell_C=\frac 1 K \sum\limits_{k=1}^K \loss{C^{\{\setminus k\}},C_{\sf 0}^{\{k\}}}
\end{equation}
In this formulation, $\ell_C$ can be recognized as the cross-validated Gaussian log-likelihood (up to a constant offset and multiplier). However, the present framework is more general, allowing for other loss functions to be used.

>>>>

\DeclareMathOperator{\Tr}{tr}
\newcommand{\sq}[1]{\lq#1\rq}
\newcommand{\mcond}{\,\middle\vert\,}
\newcommand{\cond}{\,\vert\,}
\newcommand{\figref}[2]{Fig.\;\ref{fig:#1}\,#2}
\newcommand{\loss}[1]{\mathcal L\left(#1\right)} 
\newcommand{\eloss}[1]{\mathcal L_0\left(#1\right)}
\newcommand{\T}{{\sf T}}
\newcommand{\E}[2][]{\mathbb E_{#1}\left[ #2\right]}    % expected value
\newcommand{\ie}{\emph{i.e.}\;}
\newcommand{\eg}{\emph{e.g.}\;}

@@

\section*{Abstract}
% Please keep the abstract between 250 and 300 words
Ambitious projects aim to record the activity of ever larger and denser subsets of neurons \emph{in vivo}.  Correlations in neural activity measured in such recordings can reveal important aspects of the organization of neural circuits.  However, estimating and interpreting large correlation matrices is challenging.  Estimation can be improved by regularization, \ie by imposing a structure on the estimate.  The amount of improvement depends on how closely the assumed structure represents dependencies in the data. Therefore, the selection of the most efficient correlation matrix estimator for a given neural circuit must be determined empirically.  Importantly, the identity and structure of the most efficient estimator informs about the types of dominant dependencies governing the system.

@@

Pearson correlations between the spiking activity of pairs of neurons, are among the most familiar measures of functional connectivity \cite{Averbeck:2006, Zohary:1994, Kohn:2005, Bair:2001, Ecker:2010}.  In particular, \emph{noise correlations}, \ie the correlations of trial-to-trial response variability between pairs of neurons, have a profound impact on stimulus coding \cite{Zohary:1994, Abbott:1999, Sompolinsky:2001, Nirenberg:2003, Averbeck:2006, Josic:2009, Berens:2011, Ecker:2011}. In addition, noise correlations and correlations in spontaneous activity have been hypothesized to reflect key features of the anatomical connectivity \cite{Gerstein:1964}.  This interpretation is supported by a series of discoveries of nontrivial relationships between neural correlations and other aspects of circuit organization such as the physical distances between neurons \cite{Smith:2008, Denman:2013}, their synaptic connectivity \cite{Ko:2011},  stimulus response similarity \cite{Bair:2001, Arieli:1995, Chiu:2002, Kenet:2003, Kohn:2005, Cohen:2008, Cohen:2009, Ecker:2010, Rothschild:2010, Ko:2011, Smith:2013b}, cell types \cite{Hofer:2011}, cortical layer specificity \cite{Hansen:2012, Smith:2013}, progressive changes in development and in learning \cite{Golshani:2009, Gu:2011}, changes due to sensory stimulation and global brain states \cite{Greenberg:2008, Goard:2009, Kohn:2009, Rothschild:2010, Ecker:2010, Renart:2010} among others.

Neural correlations do not come with ready or unambiguous mechanistic interpretations. They can arise from monosynaptic or polysynaptic interactions, common or correlated inputs, oscillations, top-down modulation, and background network fluctuations \cite{Perkel:1967, Moore:1970, Shadlen:1998, Salinas:2001, Ostojic:2009, Rosenbaum:2011}. But multineuronal recordings do provide more information than available from an equivalent number of pairs of cells considered separately. For example, the eigenvalue decomposition of the covariance matrix expresses shared correlated activity components across the population; common fluctuations of population activity may be accurately represented by just a few principal components but will affect all correlation coefficients. In contrast, the off-diagonal elements of the precision matrix (inverse of the covariance matrix) constitute scaled partial correlations between neuron pairs.  The partial correlation coefficient between two neurons reflects their linear association conditioned on the activity of all the other recorded cells.  For this reason, partial correlations have been proposed to more closely reflect the direct effects between components of complex systems and have been used to describe gene interactions \cite{Schafer:2005, Peng:2009} and the interactions between brain regions \cite{Varoquaux:2012, Ryali:2012}.

However, estimation of covariance matrices from large populations presents a number of numerical challenges:  The amount of recorded data grows only linearly with population size whereas the number of estimated coefficients increases quadratically.  This mismatch leads to an increase in spurious correlations, overestimation of common activity (\ie overestimation of large eigenvalues) \cite{Ledoit:2004}, and poorly conditioned partial correlations \cite{Schafer:2005}. 
The \emph{sample correlation matrix}, which is an unbiased estimate, is often used to estimate the true correlation matrix. It has many free parameters making it highly sensitive to sampling noise. As a result, on average, the sample correlation matrix is farther from the true correlation matrix than some structured estimates.

@@

To address these questions, in addition to the sample covariance matrix, we evaluated four regularized covariance matrix estimators that imposed different kinds of structure on the estimate. The estimators are designated as follows:
\begin{description}
\item[$C_{\sf 0}$] -- sample covariance matrix, the unbiased, unregularized estimator.
\item[$C_{\sf diag}$] -- linear shrinkage toward a diagonal covariance matrix. 
\item[$C_{\sf factor}$] -- linear shrinkage toward a low-rank matrix produced by factor analysis.
\item[$C_{\sf sparse}$] --  sparse partial correlations, \ie covariance selection.
\item[$C_{\sf sparse+latent}$] -- sparse partial correlations between the recorded neurons combined with partial correlations with several inferred latent units.
\end{description}

@@

\section*{Results}
% Results and Discussion can be combined.
\paragraph{Covariance estimation}
The covariance matrix is defined as
\begin{equation}\label{eq:true-covariance}
    \Sigma = \E{(x-\mu)(x-\mu)^\T},\quad \mu = \E{x}
\end{equation}
where $\E{\cdot}$ denotes expectation; the $p\times 1$ vector $x$ is a single observation the firing rates of $p$ neurons over time $\Delta t$; and $\mu$ is the vector of expected firing rates.  The usual way to estimate the covariance matrix is to calculate the \emph{sample covariance matrix} $C_{\sf 0}$ from the empirical sample of observations $x(t),\; t=n(1,\ldots,n)$ as
\begin{equation}
    C_{\sf 0} = \frac 1 \nu \sum\limits_{t=1}^n (x(t)-\bar x)(x(t)-\bar x)^\T,\quad \bar x= \frac 1 n \sum\limits_{t=1}^n x(t)
\end{equation}
where $\nu$ is the number of degrees of freedom in one neuron’s activity within the sample ($\nu=n-1$ if observations are independent). In this study, we estimate the true mean, $\mu$, using the sample mean $\bar x$, but seek a better estimate of the covariance matrix than $C_{\sf 0}$.  In the case of noise correlations, the true mean $\mu$ and the sample mean $\bar x$ are conditioned on the stimulus.

Given an arbitrary covariance matrix estimate $C$, the corresponding correlation matrix $R$ is calculated by normalizing $C$ by its diagonal (variance estimates):
\begin{equation}\label{eq:precision}
    R = \left(I\circ C\right)^{-\frac 1 2} C \left(I\circ C\right)^{-\frac 1 2},
\end{equation}
where $\circ$ denotes entrywise matrix product (Hadamard product) and $I$ is the $p\times p$ identity matrix. 
Similarly, the matrix of partial correlations $P$ is computed by normalizing the \emph{precision matrix} $C^{-1}$:
\begin{equation}\label{eq:partial}
    P = -\left(I\circ C^{-1}\right)^{-\frac 1 2} C^{-1} \left(I\circ C^{-1}\right)^{-\frac 1 2}
\end{equation}

We considered four regularized estimators based on distinct families of reduced target estimates: $C_{\sf diag}$, $C_{\sf factor}$, $C_{\sf sparse}$, and $C_{\sf sparse+latent}$. In probabilistic models with exclusively linear dependencies, the targets estimates of these for estimators can correspond to distinct families of graphical models (\figref{02}{\,Row}{row 1}).

\begin{FPfigure}
    \begin{center}
        \includegraphics{./figures/Figure02.pdf}
    \end{center}
    \caption{{\bf Estimators whose regularization targets represent the structure of the true covariance matrix outperform other estimators.}
        {\bf Row 1.} Graphical representations of the target estimates used by the four respective regularized covariance matrix estimators.  Recorded neurons are denotes as green spheres and latent units as light-shaded spheres.  The edges between them denote non-zero partial correlations.
        In estimator $C_{\sf diag}$, the regularization target is a diagonal matrix.  This target corresponds to a model wherein observed neurons exert no linear effects on one another ({\bf Row 1, A}). 
        In estimator $C_{\sf factor}$, the regularization target is a factor model (low-rank matrix plus a diagonal matrix).  This target describes a model wherein the activity of observed neurons is linearly driven by several latent units ({\bf  Row 1, B}).
        In estimator $C_{\sf sparse}$, the covariance matrix is approximated as the inverse of a sparse matrix. This approximation describes a model wherein correlations arise due to linear effects exerted by the recorded cells onto one another other ({\bf  Row 1, C}).
        In estimator $C_{\sf sparse+latent}$, the covariance matrix is approximated as the inverse of a matrix constructed as the sum of a sparse and a low-rank component. This approximation describes a model wherein correlations arise due to linear affects exerted by the recorded cells onto one another and due to linear effects from several latent units ({\bf  Row 1, D}).
\\
{\bf Row 2:} Examples of $50\times 50$ correlation matrices corresponding to each structure: the diagonal correlation matrix ({\bf  A}), a factor model with three latent units ({\bf B}),  a correlation matrix with 73\%  off-diagonal zeros ({\bf  C}), and a correlation matrix whose inverse is the some a rank-1 (\ie one latent unit) component and a sparse component with 78\%[1] off-diagonal zeros. 
\\
{\bf Row 3:} Sample correlation matrices calculated from samples of 1000 observations drawn from simulated random processes with respective correlation matrices from row 2.  The structure of the correlation matrix is much more difficult to discern by eye.
\\
{\bf Row 4:} Regularized correlation matrix estimates computed from the same samples as in Row 3.  For each kind of ground truth, the estimator with the matching structure was used and the regularization parameters optimized by cross-validation using the data in the same sample.   The regularized estimates are closer to truth than sample correlation matrices.
\\
{\bf Row 5:} Excess losses (Eq.~\ref{eq:excess-loss}) attained by each of the five estimators for  various sample sizes. The error bars indicate the standard error of the mean based on 30 samples.  Estimators with the matching structure converged to zero faster than the other estimators when the sample size was large enough to allow the more complex true structures outperform simpler structures.
\\
{\bf Row 6:} Relative cross-validation losses (Eq.~\ref{eq:rel-cv-loss}) attained  by each of the five estimators with respect to the matching estimator. Error bars indicate the standard error of the mean based on 30 samples.
    }
    \label{fig:02}
\end{FPfigure}

The target estimate of estimator $C_{\sf diag}$ is the diagonal matrix $D$ containing estimates of neurons’ variances. Regularization is achieved by linear \emph{shrinkage} of the unbiased estimate $C_{\sf 0}$ toward $D$ controlled by the scalar \emph{shrinkage intensity} parameter $\lambda \in [0, 1]$:
\begin{equation}\label{eq:c-diag}
C_{\sf diag} = (1-\lambda) C_{\sf 0} + \lambda D
\end{equation}
The structure of imposed by $C_{\sf diag}$ favors populations with no linear associations between the neurons (\figref{02}{Row 1, A}).  If sample correlations arise from spurious or contingent effects, $C_{\sf diag}$ is expected to get closer to truth than other estimators.

The target of estimator $C_{\sf factor}$ is the factor model $L + \Psi$, where $L$ is the $p\times p$ positive definite matrix with low rank and $\Psi$ is the diagonal matrix of independent variances.  The estimator, 
\begin{equation}\label{eq:c-factor}
C_{\sf factor} = (1-\lambda) C_{\sf 0} + \lambda (L + \Psi),
\end{equation}
has two hyperparameters: the number of factors in $L$ and the shrinkage intensity $\lambda$. The structure imposed by $C_{\sf factor}$ favors conditions in which the population activity is linearly driven by a number of latent factors that affect many cells while direct interactions between cells are insignificant (\figref{02}{Row 1, B}).

Estimator $C_{\sf sparse}$ is produced by approximating the sample covariance matrix by a matrix whose inverse matrix is sparse: 
\begin{equation}\label{eq:c-sparse}
C_{\sf sparse} = S^{-1},
\end{equation}
where $S$ is a sparse matrix with a large fraction of zeros in its off-diagonal.  The estimator has one hyperparameter that determines the sparsity (fraction of off-diagonal zeros) in $S$.  The structure imposed by $C_{\sf sparse}$ favors conditions in which neural correlations arise from direct linear effects (\sq{interactions}) between some pairs of neurons (\figref{02}{Row 1, C}).  We will sometimes refer to sparse partial interactions between neurons as simply \sq{interactions}.

Estimator $C_{\sf sparse+latent}$ is produced by approximating the sample covariance matrix by a matrix whose inverse is the sum of a sparse component and a low-rank component: 
\begin{equation}\label{eq:c-sl}
C_{\sf sparse+latent} = (S - L)^{-1},
\end{equation}
where, as above, $S$ is a sparse matrix and $L$ is a low rank matrix. The estimator has two hyperparameters: one to regulate the sparsity of $S$ and the other to regulate the rank of $L$. The structure imposed by $C_{\sf sparse+latent}$ favors conditions in which the activity of neurons is determined by linear effects between some observed pairs of neurons and linear effects from several latent units (\figref{02}{Row 1, D}) \cite{Chandrasekaran:2010,Ma:2013}.

\paragraph{Simulation}
To demonstrate how the selection between regularized estimators allows discerning the structure of correlation, we constructed four $50\times 50$ covariance matrices with the structures imposed by the four respective regularized estimators (\figref{02}{\,Row 2, A--D}).  See the Methods section for details of how these matrices were generated.  We used these covariance matrices as the ground truth in multivariate normal distributions with zero means and drew samples of various sizes. Even with relatively large sample sizes (\eg $n=1000$), the sample correlation matrices were contaminated by sampling noise and their underlying structure was difficult to discern (\figref{02}{\,Row 3}).

An evaluation of a covariance matrix estimators requires a \emph{loss function} $\loss{C,\Sigma}$ to quantify the discrepancy between the estimate, $C$, and the truth, $\Sigma$.  The loss function must attain its minimum when $C=\Sigma$.

In this study, we adopted the \emph{normal loss} function defined as
\begin{equation}\label{eq:loss}
    \loss{C,\Sigma} = \frac 1 p\left[\ln \det C + \Tr(C^{-1}\Sigma)\right]
\end{equation}

This loss function derives from the multivariate normal log-likelihood function $L(\Sigma|C_{\sf 0})$ through the following expression:
\begin{equation}
     \loss{\Sigma,C_{\sf 0}} = -\ln(2\pi) - \frac 2 p L(\Sigma|C_{\sf 0})
\end{equation}
The normalization by $p$ makes the values of the loss function comparable across different population sizes.

Although the two functions appear similar in form, they differ conceptually: The log-likelihood $L(\Sigma|C_{\sf 0})$ is a function of the parameter $\Sigma$ given the sample covariance matrix $C_{\sf 0}$ whereas the loss function expresses the deviation of an arbitrary estimate $C$ from the truth $\Sigma$.

The choice of normal loss function is motivated by mathematical convenience. We expect that the main conclusions of our study will not change qualitatively with other well behaved loss functions, such as Stein's entropy loss or quadratic loss \cite{James:1961, Fan:2008, Ledoit:2004, Schafer:2005}.

The \emph{excess loss} function is produced by shifting the loss function to zero at its minimum:
\begin{equation}\label{eq:excess-loss}
    \eloss{C,\Sigma} = \loss{C,\Sigma}-\loss{\Sigma,\Sigma},
\end{equation}

We drew 30 independent samples from each model of size $n=250$, 500, 1000, 2000, and 4000 and measured the excess loss produced by each of the five estimators.  The hyperparameters of the regularized estimators were optimized by nested cross-validation using only the data in the sample.  All the regularized estimators produced better estimates (lower excess loss) than the sample covariance matrix.  However, estimators whose structure matched the true model typically outperformed the other estimators (\figref{02}{\,row 5}).  There were two exceptions: First, for some smaller samples, the data were insufficient to reveal the true correlation structure and the estimators with simpler structures, $C_{\sf diag}$ and $C_{\sf factor}$ often outperformed the estimators with matching structure (\figref{02}{\,Row 5, C and D}).  Second, $C_{\sf sparse+latent}$ often performed equally well to $C_{\sf sparse}$ on the model without latent units. This can be explained by the fact that $C_{\sf sparse+latent}$ correctly inferred the right number of latent units, zero.

Even though, with increasing sample sizes, all estimators converged to the truth (zero excess loss), the distinction between the estimators did not decrease.

In empirical data the ground truth, $\Sigma$, is not accessible and excess loss cannot be computed directly. Instead, the loss function can be estimated solely from the data through \emph{validation}.  In validation, an independent \emph{testing sample} is set aside to compute an additional sample covariance estimate $C_{\sf 0}^\prime$ to validate the estimate $C$ computed from the main, \emph{training sample}.  Then \emph{validation loss} $\loss{C,C_{\sf 0}^\prime}$ can substitute the loss function $\loss{C,\Sigma}$.

@@

In this case, the cross-validation loss is an unbiased estimate of the loss function:
\begin{equation*}
    \E[C_{\sf 0}^\prime]{\loss{C,C_{\sf 0}^\prime}}=\loss{C,\E[C_{\sf 0}^\prime]{C_{\sf 0}^\prime}}=\loss{C,\Sigma}
\end{equation*}

The property of additivity does not hold for other popular loss functions such as Stein's entropy loss or various quadratic losses; their validation losses are biased estimates of the loss function.

Acquiring a separate testing sample is not practical. Instead, $K$-fold cross-validation can be used: The sample is divided into $K$ subsets of approximately equal size ($K=10$ in this study).  Then each subset is used as the validation sample with the other $K-1$ serving as the training sample. The validation losses from each of such \sq{folds} are averaged to produce the \emph{cross-validation loss} or \emph{CV-loss} for short.  Let $C_{\sf 0}^{\{k\}}$ denote the sample covariance matrix computed from the $k^{th}$ subset and $C^{\{\setminus k\}}$ denote the results of estimator $C$ trained on the remaining $K-1$ subsets. Then cross-validation loss for estimator $C$ is
\begin{equation}\label{eq:cv-loss}
    \ell_C=\frac 1 K \sum\limits_{k=1}^K \loss{C^{\{\setminus k\}},C_{\sf 0}^{\{k\}}}
\end{equation}
In this formulation, $\ell_C$ can be recognized as the cross-validated Gaussian log-likelihood (up to a constant offset and multiplier). However, the present framework is more general, allowing for other loss functions to be used.

Since we only need to compare estimators to each other, we are only interested in the \emph{relative} CV loss of estimator $C$ with respect to reference estimator $C_{\sf ref}$:
\begin{equation}\label{eq:rel-cv-loss}
    \Delta\ell_{C,C_{\sf ref}} = \ell_{C}-\ell_{C_{\sf ref}}
\end{equation}
